Inferring the existence and direction of causal associations in the face of measurement error

### 2016-12-29 docs/manuscript.Rmd
updated

The last decade has seem a rapid acceleration in technological advances that enable us to assess the human phenome in ever increasing detail, furnishing biologists with the tantalising opportunity to map the molecular basis of complex traits. This does, however, require overcoming the statistical challenge of correctly defining the causal architectures that underlie these high dimensional data. For example, identifying metabolic perturbations that lead to disease [@Kettunen2012], or understanding the influence of epigenomic influences on gene expression [@Bird2002] all depend on understanding the causal nature of any putative observational associations.

@@

Widely available observational study data are notoriously susceptible to pitfalls in making causal inference, with basic regression-based techniques unable to distinguish a true causal association from reverse causation or confounding [@DaveySmith2004; @Phillips1991; @DaveySmith2002]. In response to this, genetic instrumentation has emerged as a valuable technique for improving the reliability of causal inference in observational data, and with the coincident rise in genome-wide association studies, it is now a prominent tool that is applied in several different guises, and has been applied many forms of high throughput 'omic variables. However, an often neglected pitfall to these approaches is the influence of non-differential measurement error on the reliability of these methods.

@@

Measurement error will of course arise in more low-tech data too, for example when measuring BMI one is typically interested in using this as a proxy for adiposity, but it is clear that the correlation between BMI and underlying adiposity is not perfect [@Ahima2013]. A similar problem of biological misspecification is unavoidable in disease diagnosis, and measuring behaviour such as smoking or diet is notoriously difficult to do accurately. Measurement error can also be introduced after the data has been collected, for example the transformation of non-normal data for the purpose of statistical analysis will lead to a new variable that will typically incur both changes in scale and imprecision (noise) compared to the original variable. The sources of measurement error are not limited to this list [@Hernan2009], and its impact has been explored in the context of mediation analysis in the epidemiologyical literature extensively [@Fewell2007; @LeCessie2012; @Blakely2013].

@@

Mediation-based approaches employ genetic instruments (typically single nucleotide polymorphisms, SNPs) to orient the causal direction between the exposure and the outcome. If a SNP is associated with an exposure, and the exposure is associated with some outcome, then it logically follows that the implied influence of the SNP on the outcome will disappear when conditioning on the exposure. Here, the exposure 'mediates' the association between the SNP and the outcome, providing information about the causal influence of the exposure on the outcome. This forms the basis of a number of methods such as genetical genomics [@Schadt2005], the regression-based causal inference test (CIT) [@Millstein2009; @Millstein2016], an SEM implementation in the NEO software [@Aten2008], and various other methods including Bayesian approaches [@Waszak2015]. They have been employed by a number of recent publications that make causal inferences, often in large scale ‘omics datasets [@Koestler2014; @Liu2013; @Waszak2015; @Yuan2014; @Tang2014; @Hong2015].

@@

By utilising genetic instruments in different ways, mediation-based analysis and MR models have properties that confer some advantages and some disadvantages. The main advantage of mediation over MR is that the true underlying biology of the genetic instrument doesn't necessarily need to be known as long as it associates with the putative exposure. In the CIT framework (described fully in the Methods) for example, the test statistic is different if you test for the exposure causing the outcome or the outcome causing the exposure, allowing the researcher to infer the direction of causality between two variables by performing the test in both directions and choosing the model with the strongest evidence.

@@

MR does however have some important advantages over mediation based approaches. First, MR is used as a tool to not just deduce the existance of causality, but to also estimate the causal effect size. Second, using mediation requires that the exposure, outcome and instrument variables are all measured in the same data, whereas recent extensions to MR circumvent this requirement, allowing causal inference to be drawn when exposure variables and outcome variables are measured in different samples [@Pierce2013]. This has the crucial advantage of improving statistical power by allowing analysis in much larger sample sizes. Third, and though this is often underappreciated in fields outside classical epidemiology it is of crucial importance, MR is substantially more robust to there being measurement error in the exposure and outcome variables. Indeed instrumental variable (IV) analysis was in part initially introduced as a correction for measurement error [@Ashenfelter1994], whereas it has been noted that both basic mediation [@LeCessie2012; @Nagarajan2013; @Shpitser2010; @Blakely2013] and mediation methods that use instrumental variables [@Wang2015; @Lagani2015] are prone to be unreliable in its presence. We argue that this is a serious limitation.

Here we show using theory and simulations how non-differential measurement error and measurement imprecision can lead to unreliable causal inference in the mediation-based CIT method. We also present an extension to MR that allows researchers to ascertain the causal direction even when the biology of the instrument is not fully understood. This improves the utility of MR in cases where mediation based methods might have otherwise been used preferentially, and we apply this new method to infer the direction of causation between DNA methylation levels and gene expression levels.

>>>>

@@

Observational measures of the human phenome are growing ever more abundant, but using these data to make causal inference is notoriously susceptible to many pitfalls, with basic regression-based techniques unable to distinguish a true causal association from reverse causation or confounding [@Phillips1991; @DaveySmith2002; @DaveySmith2004]. In response to this, genetic instrumentation has emerged as a technique for improving the reliability of causal inference in observational data, and with the coincident rise in genome-wide association studies it is now a prominent tool that is applied in several different guises [@DaveySmith2004; @Millstein2009; @Aten2008; @Waszak2015]. However, potential pitfalls still persist and one that is often neglected is the influence of non-differential measurement error on the reliability of causal inference.

@@

Measurement error will of course arise in more low-tech data too, for example when measuring BMI one is typically interested in using this as a proxy for adiposity, but it is clear that the correlation between BMI and underlying adiposity is not perfect [@Ahima2013]. A similar problem of biological misspecification is unavoidable in disease diagnosis, and measuring behaviour such as smoking or diet is notoriously difficult to do accurately. Measurement error can also be introduced after the data has been collected, for example the transformation of non-normal data for the purpose of statistical analysis will lead to a new variable that will typically incur both changes in scale and imprecision (noise) compared to the original variable. The sources of measurement error are not limited to this list [@Hernan2009], and its impact has been explored in the epidemiological literature extensively [@Fewell2007; @LeCessie2012; @Blakely2013].

@@

Mediation-based approaches employ genetic instruments (typically single nucleotide polymorphisms, SNPs) to orient the causal direction between the exposure and the outcome. If a SNP is associated with an exposure, and the exposure is associated with some outcome, then it logically follows that the indirect influence of the SNP on the outcome will disappear when conditioning on the exposure. Here, the exposure 'mediates' the association between the SNP and the outcome, providing information about the causal influence of the exposure on the outcome. This forms the basis of a number of methods such as genetical genomics [@Schadt2005], the regression-based causal inference test (CIT) [@Millstein2009; @Millstein2016], a structural equation modelling (SEM) implementation in the NEO software [@Aten2008], and various other methods including Bayesian approaches [@Waszak2015]. They have been employed by a number of recent publications that make causal inferences in large scale ‘omics datasets [@Koestler2014; @Liu2013; @Waszak2015; @Yuan2014; @Tang2014; @Hong2015].

@@

By utilising genetic instruments in different ways, mediation-based analysis and MR models have properties that confer some advantages and some disadvantages for reliable causal inference. The main advantage of mediation over MR is that the true underlying biology of the genetic instrument doesn't necessarily need to be known as long as it associates with the putative exposure. In the CIT framework (described fully in the Methods) for example, the test statistic is different if you test for the exposure causing the outcome or the outcome causing the exposure, allowing the researcher to infer the direction of causality between two variables by performing the test in both directions and choosing the model with the strongest evidence.

@@

MR does however have some important advantages over mediation based approaches. First, MR is used as a tool to not just deduce the existance of causality, but to also estimate the causal effect size. Second, using mediation requires that the exposure, outcome and instrument variables are all measured in the same data, whereas recent extensions to MR circumvent this requirement, allowing causal inference to be drawn when exposure variables and outcome variables are measured in different samples [@Pierce2013]. This has the crucial advantage of improving statistical power by allowing analysis in much larger sample sizes, and dramatically expands the breadth of possible phenotypic relationships that can be evaluated. Third, MR is substantially more robust to there being measurement error in the exposure and outcome variables. Indeed instrumental variable (IV) analysis was in part initially introduced as a correction for measurement error [@Ashenfelter1994], whereas it has been noted that both classic mediation-based analyses [@LeCessie2012; @Nagarajan2013; @Shpitser2010; @Blakely2013] and mediation-based methods that use instrumental variables [@Wang2015; @Lagani2015] are prone to be unreliable in its presence. We argue that this is a serious limitation.

Here we show using theory and simulations how non-differential measurement error and measurement imprecision can lead to unreliable causal inference in the mediation-based CIT method. We then present an extension to MR that allows researchers to ascertain the causal direction of an association even when the biology of the instruments are not fully understood, and to evaluate the sensitivity of the result to measurement error. This improves the utility of MR in cases where mediation based methods might have otherwise been used preferentially. Finally, we apply this new method to infer the direction of causation between DNA methylation levels and gene expression levels.
